---
title: "Regression Models Project 1"
author: "Dann Hekman"
date: "Saturday, March 07, 2015"
output: html_document
---

##Evaluating Fuel Economy of Manual vs. Automatic Transmissions using Regression Analysis##

###Executive Summary###  
There is a common belief that cars with a manual transmission are more fuel-efficient than automatics. Using 1974 [data](https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/mtcars.html) from *Motor Trends* we can analyze this claim using linear regression analysis. Based on the analysis presented below, although manual transmission cars get on average 7.24 more miles per gallon than automatic cars, we can confidently say that this is because automatic transmission vehicles weigh more, and therefore have lower gas mileage. Based on these data, we do not believe that manual transmission vehicles have a better gas mileage, once you account for the weight of the vehicle. 

###Research Question###
We start our analysis by examining the claim that manual transmission cars have better gas mileage than automatic cars. We can load the data and do some basic analysis in R on this dataset. 
```{r Short Model A}
data(mtcars)
str(mtcars)
summary(mtcars)
```  
Specifically, our variables of interest are mpg and am, which is coded to 0 for automatic transmissions, and 1 for manual transmission. 
We can regress  
$$\tiny(1) \normalsize \hspace{1cm} Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$$  
where $X_i$ is the *as* variable in the mtcars dataset. Simple Least Squares regression will give the average mpg for automatics ($X = 0$) and $\beta_1$ will give the added average miles per gallon of a manual transmission. 
Looking at just the transmission, we can see that automatic transmission vehicles have an average of 17.15 miles per gallon, and manuals get an additional 7.24 miles per gallon on average (24.39 total).

```{r Short Model B}
model <- lm(mpg ~ am, data=mtcars)
summary(model)$coef
```  

###Correction for Car Weight###
Before jumping to the conclusion that manual transmissions are more fuel efficient, we should try to control for other variables in the dataset. Specifically, let's look at the number of cylinders, horsepower, and vehicle weight.  
$$\tiny(2) \normalsize \hspace{1cm} Y_i = \beta_0 + \beta_1 X_i + \sum \beta_j X_ij + \epsilon_i$$  
where $\beta_1$ is the coefficient for the dummy variable *as* and the sum of $\beta_j$ and $X_j$ includes the *cyl, hp,* and *wt* variables from the data frame.  
```{r Long Model}
model <- lm(mpg ~ am + cyl + hp + wt, data=mtcars)
summary(model)$coef
```  

Controlling for weight, manual vs. automatic ceases to be a statistically significant predictor of gas mileage

###Conclusions###
Despite the fact that manual transmission vehicles do, on average, have a better gas mileage, when controling for the number of cylinders, horsepower, and most importantly vehicle weight, this difference ceases to be substantive or significant. This spurious relationship between mpg, transmission, is very well illustrated by the following graph.

```{r Graph}
library(ggplot2)
q<-qplot(x=wt,y=mpg,data=mtcars,colour=as.factor(am))
q<-q + scale_color_discrete(name="Transmission",labels=c("Automatic","Manual"))
q<-q + ggtitle("Miles per Gallon by Weight and Transmission")
q<-q + xlab("Weight (1000lbs)") + ylab("Miles Per Gallon")
q
```  

So we can see clearly that automatics tend to be both heavier and have a lower gas mileage, but comparibly weighted manual transmission vehicles tend to have comparible gas mileage. The 7.24 miles-per-gallon difference that we saw on average between manual and automatic vehicles in figure (1) was primarily due not the the better fuel efficiency of automatic transmissions, but the fact that most automatic transmission cars also tend to weigh less than automatics and have less horsepower, as seen in equation (2).
